---
title: "augmentation"
author: 
  - Rune Daucke 
  - David Faurdal
  - Luisa Weisch
editor: visual
---

## Load libraries and data

```{r}
#| label: load_libraries_augment
#| eval: TRUE
#| echo: TRUE
#| results: hide
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(viridis)
source("99_proj_func.R")
```

```{r}
#| label: load_tibble
#| eval: TRUE
#| echo: TRUE
#| results: hide
#| message: FALSE
#| warning: FALSE
df <- read_tsv("../data/02_dat_clean.tsv")
all_genes <- colnames(df) %>% .[- c(1,2,3)]

```

## Calculating the Mean Expression for Specific Gene Sets

### Neurodegernerative gene set

Calculating the mean of the expression of genes which are part of neurodegeneration. Gene list from: https://pmc.ncbi.nlm.nih.gov/articles/PMC3044594/#:\~:text=Mutations%20in%206%20genes%20(SNCA,shown%20to%20cause%20familial%20parkinsonism.

```{r}
#| label: Extract values for ndg genes
#| eval: TRUE
#| echo: TRUE
#| results: hide
#| message: FALSE
#| warning: FALSE
# Neurodegenerative-associated genes
ndg_list <- c("SNCA", "LRRK2", "PRKN", "DJ1", "PINK1", "ATP13A2")

present_genes <- find_genes(ndg_list, all_genes)
df$ndg_mean <- rowMeans(df[present_genes], na.rm = TRUE)

# print(c("Number of present genes:", length(present_genes)))
# df$ndg_mean
```

### Inflamation gene set

Next we look at another set of genes which play a role in inflammation. The aim here is to determine whether inflammation is higher in patients versus controls. Gene list from: https://pmc.ncbi.nlm.nih.gov/articles/PMC4955784/#:\~:text=Through%20gene%20expression%20microarray%20analysis,%2C%20Il1rn%2C%20Saa3%2C%20Emr1%2C

```{r}
#| label: Extract values for inflammatory genes
#| eval: TRUE
#| echo: TRUE
#| results: hide
#| message: FALSE
#| warning: FALSE
#| 
# inflammation index
#inflam_list <- c("IL6", "TNF", "IL1B", "CXCL10")
inflam_list <- c("IL1rn", "TNF", "Saa3", "Emr1", "Adam8", "Itgam")

present_genes <- find_genes(inflam_list, all_genes)
df$inflam_mean <- rowMeans(df[present_genes], na.rm = TRUE)

```

# Saving augmented .tsv file

```{r}
write_tsv(x = df, file="../data/03_dat_aug.tsv")
```
