---
title: "Analysis 1 - Expression in Gene Sets"
author: 
  - Rune Daucke 
  - David Faurdal
  - Luisa Weisch
format: html
editor: visual
---

# Expression in Gene Sets

## Load Libraries and Augmented Data

```{r}
#| label: load_libraries_gene_sets
#| eval: TRUE
#| echo: TRUE
#| results: hide
#| message: FALSE
#| warning: FALSE

library(tidyverse)
#library(hrbrthemes)
library(viridis)
library(ggplot2)
```

```{r}
#| label: load_data_gene_sets
#| eval: TRUE
#| echo: TRUE
#| results: hide
#| message: FALSE
#| warning: FALSE
df <- read_tsv("../data/03_dat_aug.tsv")

```

In 03_augmentation.qmd, we calculated the mean gene expression for set of genes. In the following code we will analyse the results

## Visualize

### Different Expression of Gene Sets

In 03_augment.qmd we calculated the mean of some gene sets. \#### Distribution and Boxplot

In the following chunks we compare the expression levels using density plots and boxplots. \##### Neurodegenerative Gene Set

```{r}
#| label: plot_of_neurodegenerative_genes
#| eval: TRUE
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
#| fig.width: 8
#| fig.height: 6

# Plotting distribution
ggplot(data = df, 
       aes(x = ndg_mean, 
           fill = Condition)) +
  geom_density(alpha = 0.5) +  # alpha controls transparency
  labs(title = "Density Plot for Neurodegenerative Genes",
       x = "Gene Expression",
       y = "Density") +
  scale_fill_manual(values = c("Control" = "blue", "PD" = "red")) +
  theme_minimal()

ggsave("../results/07_neurodeg_density.png", width = 8, height = 6, dpi = 600)


df %>% 
  ggplot(aes(x=Condition, y=ndg_mean, fill=Condition)) +
  geom_boxplot() + 
  scale_fill_viridis(discrete = TRUE, alpha=0.6) +
  labs(y = "Mean Gene Expression") +
  theme(
    legend.position="none",
    plot.title = element_text(size=11)
  ) +
  ggtitle("Expression of neurodegenerative-associated genes, comparing control with disease") +
  xlab("") + 
  theme_minimal()

ggsave("../results/07_neurodeg_boxplot.png", width = 8, height = 6, dpi = 600)

```

##### Pro-inflammatory cytokines and chemokines

```{r}
#| label: plot_of_proinflammatory_genes
#| eval: TRUE
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
#| fig.width: 8
#| fig.height: 6


# Plotting distribution
ggplot(df, aes(x = inflam_mean, fill = Condition)) +
  geom_density(alpha = 0.5) +  # alpha controls transparency
  labs(title = "Density Plot for Pro-Inflamatory  Cytokines",
       x = "Gene Expression", y = "Density") +
  scale_fill_manual(values = c("Control" = "blue", "PD" = "red")) +
  theme_minimal()

ggsave("../results/07_inflammatory_density.png", width = 8, height = 6, dpi = 600)

# Boxplot
df %>% 
  ggplot(aes(x=Condition, y=inflam_mean, fill=Condition)) +
  geom_boxplot() + 
  scale_fill_viridis(discrete = TRUE, alpha=0.6) +
  labs(y = "Mean Gene Expression") +
  theme(
    legend.position="none",
    plot.title = element_text(size=11)
  ) +
  ggtitle("Mean Expression of Inflammation-related genes, comparing control and disease") +
  xlab("") + 
  theme_minimal()

ggsave("../results/07_inflammatory_boxplot.png", width = 8, height = 6, dpi = 600)
```

## Interpretation - Statistical relevance

In the first boxplot, showing the different mean expression level for neurodegenerative-associated genes, a slight difference is visible. The second boxplot already shows less difference. The plots are a good and quick way to understand the data better, but if there is a significant difference is hard to see, so in the next chunk we look at the statistical relevance of the difference, by calculating the p-value.

```{r}
t_test_result <- t.test(ndg_mean ~ Condition, data = df)

print(t_test_result)
```

```{r}
t_test_result <- t.test(inflam_mean ~ Condition, data = df)

print(t_test_result)
```

Both test showed us that there is no statistical relevance, with the p value being even higher then 0.3, and 0 being included in the 95% confidence interval.
